A Study on Common Bug Fix Patterns in Rust

### 2022-10-14 sections/methodology.tex
added some minor changes, about to fix results

    \caption{\label{fig:extraction}Extracting paths from an ASTDiff. the ASTDiff on left corresponds to the code change in the yellow box on the top right. A DFS search will identify three paths, shown in three different colors at right.}

@@

        \node[draw, minimum height=0.5, fill=yellow!20] at (4.02,-0.25) {Weights data};

@@

    \caption{\label{fig:essence}Multiplying the number of occurrences of items observed in the tree by their respective weights, resulting in the essence of change (visualized with a circle pack figure at right). The left and right key value tables show the number of occurrences of each item and its respective weight, respectively.}
\end{figure}

>>>>

    \caption{\label{fig:extraction}The ASTDiff (left) corresponds to the code change in the yellow box (top right). A DFS will identify three paths, shown in three different colors at right.}

@@

        \node[draw, minimum height=0.5, fill=yellow!20] at (4.12,-0.25) {Weighted data};

@@

    \caption{\label{fig:essence}Multiplying the number of occurrences of items observed in the tree by their respective weights, resulting in the essence of change (visualized with a circle pack figure at bottom right). Vectors show (above) the number of occurrences and (below) weighted occurrence counts.}
\end{figure}

### 2022-10-14 sections/results.tex
added some minor changes, about to fix results

\subsection{Attributes}

@@

\noindent \textit{\label{sec:G.attr.struct}\textbf{G.attr.struct} Modifying the attributes of structs}

@@

\noindent \textbf{Description:} In this pattern, the attribute set of a struct changes. This modification can happen in three different ways: (G.attr.struct.add) adding attributes to the struct; (G.attr.struct.drop) removing attributes from the struct; and (G.attr.struct.change) changing the content of an existing attribute. These attributes are Outer Attributes as we can find struct related non-terminals within the ASTDiff of the instances of this pattern. \\

\noindent \textit{ \textbf{G.attr.struct.field} Modifying the attributes of struct fields}

@@

\noindent \textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, this modification also happens in three different ways. Using attributes for struct fields, apart from being able to store different meta-information for different struct fields, for some use cases, the programmers can control the action they want to apply on different fields.
% this last sentence immediately prior to here has too many moving parts

\subsection{Struct} 
\noindent \textit{ \textbf{G.struct.field} Adding/Removing a struct field}

@@

\noindent \textbf{Description:} A set of new fields are added to or removed from an existing struct. A developer adds a new field to store a new piece of data in a struct. A developer removes a field when the developer realizes that the field is not required with respect to the behaviour they want to implement (possibly because they have moved it elsewhere). \\

\noindent \textit{ \textbf{G.struct.field.pub} Modifying the access modifiers of struct fields}

@@

\noindent \textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in a bug fixing change, where the developer needed to access a field which had not been marked for public access. On the other hand, the public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes.

\subsection{Option}
\vspace{3mm}

\noindent \textit{ \textbf{G.option} Changing field type T to T$\langle$Option$\rangle$}

@@

\noindent \textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If we want a variable to contain null, we would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that our variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field.

@@

\subsection{Types}
\vspace{3mm}

\noindent \textit{\label{sec:G.type.field}\textbf{G.type.field} Changing a struct field type}

@@

\noindent \textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. For instance, the developer may be implementing a new feature or fixing the behaviour of the program. Also it can happen for refactoring purposes or performance enhancement. \\
% maybe discuss the specific example a bit?

\noindent \textit{ \textbf{G.type.generic} Changing a generic type parameter}

@@

\noindent \textbf{Description:} This a change of the type parameter of a struct field. As with \textit{\textbf{G.type.field}}, this change can also happen for the mentioned reasons. \\

\noindent \textit{ \textbf{G.type.enum.variant} Change in enum variant value type}

@@

\noindent \textbf{Description:} In Rust, developers can specify the type of the value that an enum variant can store. Using this feature, they can avoid having to use a struct along with enum variants. This change pattern is associated with a change in enum variant types. \\

\noindent \textit{ \textbf{G.type.tuple} Changing the type inside a tuple}

@@

\noindent \textbf{Description:} In this pattern, an extra type is added to an existing type, making it a tuple of multiple types. Alternatively, the change might be dropping types from the tuple and making it a smaller-arity tuple.

\subsection{Traits}
\noindent \textit{ \textbf{G.trait.bound} TraitBound Change}

@@

\noindent \textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This change pattern relates to modification in trait bounds. This can happen both for bug fix reasons or just refactoring purposes.

@@

\noindent \textit{ \textbf{BC.clone.drop} Removing clone}

@@

\noindent \textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy\footnote{\url{https://github.com/rust-lang/rust-clippy}} is a linter tool for Rust that catches common mistakes. There are more than 500 lints included in Clippy. With regards to this pattern, Clippy detects whether a variable clone is necessary and could we have used the variable itself or not. This change does not modify the program behaviour, however it can boost the performance significantly if we repeatedly cloning large variables.

\noindent \textit{ \textbf{BC.clone.ref} Dropping clone and adding borrowing}

@@

\noindent \textbf{Description:} Unlike the previous pattern, this pattern is not detected by Clippy. The change happens when the programmers realizes that the cloning of a variable is unnecessary, however, we also want to preserve the ownership of the variable within the current scope. That is why in this pattern a clone is turned into a borrowing.

@@

\noindent \textit{ \textbf{BC.ref.add} Adding Borrowing}

@@

\noindent \textbf{Description:} This change happens when the developer needs to modify the program so it reaches the desired behaviour. The modification, however, requires access to an object whose ownership has moved to another scope and never returned. That is why we must use borrowing to preserve ownership within our scope, after the function execution finishes.

@@

\noindent \textit{ \textbf{BC.ref.drop} Removing Borrowing}

@@

\noindent \textbf{Description:} In the pattern, an unnecessary borrowing is removed. Clippy can detect the needless borrows, and also can automatically remove them. This does not change the behaviour of the program, however, it does improve the performance, as the borrow checker need not to return the ownership after the execution of the function. \\

\noindent \textit{ \textbf{BC.deref.to\_string()} Modification in dereferencing a string prior to calling to\_string()}

@@

\noindent \textbf{Description:} This is a change that in all the instances has been detected by Clippy. The change applies on an object prior to calling \verb+to_string()+ on that object. The change might add or remove a dereference, as shown above. In the addition case, the reason behind the change is that the object a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. It is faster and more performance efficient to dereference the object and then call \verb+to_string()+ on it. The deletion case indicates a redundant dereference, thus it is removed.

@@

\noindent \textit{ \textbf{BC.mut.add} Adding mutability}

@@

\noindent \textbf{Description:} This change happens when the programmer needs to mutate a variable which is immutable, hence they need to add the mut keyword. In the observed instances, this pattern was not seen in refactoring changes, the modification happened alongside a change in program behaviour. \\

\noindent \textit{ \textbf{BC.mut.drop} Dropping mutability}

@@

\noindent \textbf{Description:} Dropping the mut keyword before a variable, as we do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed through this tool.

@@

\noindent \textit{ \textbf{BC.vec.slice} Changing a Vec reference to Slice}

\begin{lstlisting}[language=Rust, style=colouredRust]
pub fn expand_delimited_square(
--  children: &Vec<TokenNode>,
++  children: &[TokenNode],
) -> Result<hir::Expression, ParseError> {
    // body
}
\end{lstlisting}
\noindent \textbf{Description:} In struct field definition or the types of the function formal parameters, type \verb+&Vec<T>+ changes to \verb+&[T]+. Clippy can detect the opportunity to apply this pattern. The reason is reducing type complication, as slice types in the form of \verb+&[T]+ or \verb+&str+ are usually sufficient for most use cases.

\subsubsection{Lifetime}
\noindent \textit{ \textbf{BC.lifetime.static.drop} Dropping static lifetime}

@@

\noindent \textbf{Description:} Removing the static keyword from a const variable declaration. This change can be applied using Clippy. If the presence of static is not required, it is better that we omit it as keeping it might create complicate types in our program.

>>>>

\subsubsection{Attributes}

@@

\noindent\textit{\label{sec:G.attr.struct}\textbf{G.attr.struct} Modifying the attributes of structs}

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct changes. This modification can happen in three different ways: (G.attr.struct.add) adding attributes to the struct; (G.attr.struct.drop) removing attributes from the struct; and (G.attr.struct.change) changing the content of an existing attribute. These attributes are Outer Attributes as we can find struct related non-terminals within the ASTDiff of the instances of this pattern. \\

\noindent\textit{\textbf{G.attr.struct.field} Modifying the attributes of struct fields}

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, this modification also happens in three different ways. Using attributes for struct fields, apart from being able to store different meta-information for different struct fields, for some use cases, the programmers can control the action they want to apply on different fields.
% this last sentence immediately prior to here has too many moving parts

\subsubsection{Struct} 
\noindent\textit{\textbf{G.struct.field} Adding/Removing a struct field}

@@

\noindent\textbf{Description:} A set of new fields are added to or removed from an existing struct. A developer adds a new field to store a new piece of data in a struct. A developer removes a field when the developer realizes that the field is not required with respect to the behaviour they want to implement (possibly because they have moved it elsewhere). \\

\noindent\textit{\textbf{G.struct.field.pub} Modifying the access modifiers of struct fields}

@@

\noindent\textbf{Description:} Adding \texttt{pub} to a struct field makes it possible to access that field from external modules. In the instances of this cluster, adding the \texttt{pub} access modifier happened in a bug fixing change, where the developer needed to access a field which had not been marked for public access. On the other hand, the public access might be revoked if the developer realizes there are no external accesses to the field, and that such accesses are not desirable, e.g. for encapsulation purposes.

\subsubsection{Option}
\vspace{3mm}

\noindent\textit{\textbf{G.option} Changing field type T to T$\langle$Option$\rangle$}

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If we want a variable to contain null, we would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that our variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field.

@@

\subsubsection{Types}
\vspace{3mm}

\noindent\textit{\label{sec:G.type.field}\textbf{G.type.field} Changing a struct field type}

@@

\noindent\textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. For instance, the developer may be implementing a new feature or fixing the behaviour of the program. Also it can happen for refactoring purposes or performance enhancement. \\
% maybe discuss the specific example a bit?

\noindent\textit{\textbf{G.type.generic} Changing a generic type parameter}

@@

\noindent\textbf{Description:} This a change of the type parameter of a struct field. As with \textit{\textbf{G.type.field}}, this change can also happen for the mentioned reasons. \\

\noindent\textit{\textbf{G.type.enum.variant} Change in enum variant value type}

@@

\noindent\textbf{Description:} In Rust, developers can specify the type of the value that an enum variant can store. Using this feature, they can avoid having to use a struct along with enum variants. This change pattern is associated with a change in enum variant types. \\

\noindent\textit{\textbf{G.type.tuple} Changing the type inside a tuple}

@@

\noindent\textbf{Description:} In this pattern, an extra type is added to an existing type, making it a tuple of multiple types. Alternatively, the change might be dropping types from the tuple and making it a smaller-arity tuple.

\subsubsection{Traits}
\noindent\textit{\textbf{G.trait.bound} TraitBound Change}

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This change pattern relates to modification in trait bounds. This can happen both for bug fix reasons or just refactoring purposes.

@@

\noindent\textit{\textbf{BC.clone.drop} Removing clone}

@@

\noindent\textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy\footnote{\url{https://github.com/rust-lang/rust-clippy}} is a linter tool for Rust that catches common mistakes. There are more than 500 lints included in Clippy. With regards to this pattern, Clippy detects whether a variable clone is necessary and could we have used the variable itself or not. This change does not modify the program behaviour, however it can boost the performance significantly if we repeatedly cloning large variables.

\noindent\textit{\textbf{BC.clone.ref} Dropping clone and adding borrowing}

@@

\noindent\textbf{Description:} Unlike the previous pattern, this pattern is not detected by Clippy. The change happens when the programmers realizes that the cloning of a variable is unnecessary, however, we also want to preserve the ownership of the variable within the current scope. That is why in this pattern a clone is turned into a borrowing.

@@

\noindent\textit{\textbf{BC.ref.add} Adding Borrowing}

@@

\noindent\textbf{Description:} This change happens when the developer needs to modify the program so it reaches the desired behaviour. The modification, however, requires access to an object whose ownership has moved to another scope and never returned. That is why we must use borrowing to preserve ownership within our scope, after the function execution finishes.

@@

\noindent\textit{\textbf{BC.ref.drop} Removing Borrowing}

@@

\noindent\textbf{Description:} In the pattern, an unnecessary borrowing is removed. Clippy can detect the needless borrows, and also can automatically remove them. This does not change the behaviour of the program, however, it does improve the performance, as the borrow checker need not to return the ownership after the execution of the function. \\

\noindent\textit{\textbf{BC.deref.to\_string()} Modification in dereferencing a string prior to calling to\_string()}

@@

\noindent\textbf{Description:} This is a change that in all the instances has been detected by Clippy. The change applies on an object prior to calling \verb+to_string()+ on that object. The change might add or remove a dereference, as shown above. In the addition case, the reason behind the change is that the object a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. It is faster and more performance efficient to dereference the object and then call \verb+to_string()+ on it. The deletion case indicates a redundant dereference, thus it is removed.

@@

\noindent\textit{\textbf{BC.mut.add} Adding mutability}

@@

\noindent\textbf{Description:} This change happens when the programmer needs to mutate a variable which is immutable, hence they need to add the mut keyword. In the observed instances, this pattern was not seen in refactoring changes, the modification happened alongside a change in program behaviour. \\

\noindent\textit{\textbf{BC.mut.drop} Dropping mutability}

@@

\noindent\textbf{Description:} Dropping the mut keyword before a variable, as we do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed through this tool.

@@

\noindent\textit{\textbf{BC.vec.slice} Changing a Vec reference to Slice}

\begin{lstlisting}[language=Rust, style=colouredRust]
pub fn expand_delimited_square(
--  children: &Vec<TokenNode>,
++  children: &[TokenNode],
) -> Result<hir::Expression, ParseError> {
    // body
}
\end{lstlisting}
\noindent\textbf{Description:} In struct field definition or the types of the function formal parameters, type \verb+&Vec<T>+ changes to \verb+&[T]+. Clippy can detect the opportunity to apply this pattern. The reason is reducing type complication, as slice types in the form of \verb+&[T]+ or \verb+&str+ are usually sufficient for most use cases.

\subsubsection{Lifetime}
\noindent\textit{\textbf{BC.lifetime.static.drop} Dropping static lifetime}

@@

\noindent\textbf{Description:} Removing the static keyword from a const variable declaration. This change can be applied using Clippy. If the presence of static is not required, it is better that we omit it as keeping it might create complicate types in our program.
